Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-25 submission/manuscript.Rmd
move discussion notes to end
edit abstract

@@

## Discussion notes

@@

\clearpage
# Notes

>>>>

It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). While some chromatin regulators are important for neuron-specific gene expression, loss of repressive chromatin regulators can also lead to the aberrant transcription of tissue-specific genes outside of their intended context. Because very few studies have characterized the these aberrant, tissue-specific transcripts, the mechanisms behind their dysregulation and their functional consequences are largely unknown. Here, we explored this crisis in cellular identity in NDDs by investigating the ectopic expression of germline (e.g. sperm and egg) genes in mice lacking lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability (MRXSCJ, OMIM: 300534), aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Many genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary. Using germ cell-depleted RNA sequencing datasets, we curated a a list of germline-enriched genes for unbiased characterization of KDM5C in germline gene repression. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. We found that while KDM5C promotes the initial placement of DNA methylation at a subset of germline gene promoters in EpiLCs, germline genes can also become activated independent of direct KDM5C regulation. Intriguingly, many of the germline transcripts expressed in the _Kdm5c_-KO brain were late-stage spermatogenesis genes that are not directly bound by KDM5C. This suggests germline developmental programs can progress ectopically in the background of typical _Kdm5c_-KO development. Ultimately, this work provides novel insight into the demarcation of somatic and germline lineages in mammals while also characterizing the cellular identity crisis within a chromatin-linked neurodevelopmental disorder.

@@

@@

\clearpage
# Notes
## Discussion notes
